---
title: "Log response ratio example: Macartney et al. 2022"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

[Macartney, Lagisz, and Nakagawa (2022)](https://www.sciencedirect.com/science/article/abs/pii/S0149763422000434) conducted a meta-analysis examining how environmental enrichment and stress affect learning and memory in rodents. They compiled data from 30 studies that used a fully crossed 2×2 factorial design and computed log response ratios (lnRR). Here I focus on three of them: the main effect of enrichment, the main effect of stress, and their interaction. These effect sizes were then analysed using multi-level meta-analysis models.

This vignette shows how to reproduce those three core results using the `minter` package. Instead of writing custom functions (as in the original analysis), `minter` provides `lnRR_main()` and `lnRR_inter()` which compute the effect sizes and their sampling variances directly from raw group means, standard deviations, and sample sizes. 

## Loading packages

```{r setup}
library(minter)
library(orchaRd)
library(metafor)
library(ggplot2)
```

## The data

Each included study used a fully crossed 2×2 factorial design with four treatment groups:

| Group | Housing | Stress |
|---|---|---|
| CC | conventional | no |
| EC | enriched | no |
| CS | conventional | yes |
| ES | enriched | yes |

Before the analysis, I have to load the raw data and do some cleaning and processing. These steps are almost identical in the original code. First exclude the Wang study (an outlier in the original analysis) and floor the sample sizes:

```{r}
df <- read.csv(
  system.file("extdata", "Macartney2022_raw.csv", package = "minter"),
  stringsAsFactors = TRUE
)

df <- droplevels(df[!df$First_author == "Wang", ])

df$CC_n <- floor(df$CC_n)
df$EC_n <- floor(df$EC_n)
df$CS_n <- floor(df$CS_n)
df$ES_n <- floor(df$ES_n)
```

For each group the raw data reports a sample size, a mean, and a standard deviation. The original dataset also contains moderator variables (species, strain, sex, assay type, etc.). 

The dataset contains 92 effect sizes from 30 studies. Here are the key columns:

```{r}
head(df[, c("Study_ID", "First_author", "Year_published", "ES_ID",
             "CC_n", "CC_mean", "CC_SD",
             "EC_n", "EC_mean", "EC_SD",
             "CS_n", "CS_mean", "CS_SD",
             "ES_n", "ES_mean", "ES_SD",
             "Response_percent", "Response_direction")])
```

To compute the log response ratios, twelve columns are required: for each of the four groups, a mean, SD, and sample size. In this dataset they are named `CC_mean`, `CC_SD`, `CC_n`, and so on. However, before computing the effect sizes, proportion data must be transformed.

## Arcsine transformation for proportion data

Here we apply the arcsine square-root transformation to proportions (see equations 9 and 10 in [Macartney et al. 2022](https://www.sciencedirect.com/science/article/abs/pii/S0149763422000434)):

```{r}
mean_asin <- function(x) {
    asin(sqrt(x / 100))
}

sd_asin <- function(x, sd) {
    sqrt(((sd / 100)^2) / (4 * (x / 100) * (1 - (x / 100))))
}
```

Note that the formulas divide by 100 (i.e., `(x / 100)`). This is because these functions expect percentages as input. The authors included a column called `Response_percent`, so I have to transform all those rows.

Here I create columns called `t_xx_xxxx` to store the values that are going to be used in the log response ratios. In the next step, I apply the arcsine transformation to those values which are percentages:

```{r}
df[["t_CC_mean"]] <- df[["CC_mean"]]
df[["t_CC_SD"]]   <- df[["CC_SD"]]
df[["t_EC_mean"]] <- df[["EC_mean"]]
df[["t_EC_SD"]]   <- df[["EC_SD"]]
df[["t_CS_mean"]] <- df[["CS_mean"]]
df[["t_CS_SD"]]   <- df[["CS_SD"]]
df[["t_ES_mean"]] <- df[["ES_mean"]]
df[["t_ES_SD"]]   <- df[["ES_SD"]]
```

Here I use the `Response_percent` column to create a vector that filters rows with percentages. Then, apply transformation only to those rows:

```{r}
is_perc <- df[["Response_percent"]] == "yes"

df[is_perc, "t_CC_mean"] <- mean_asin(df[is_perc, "CC_mean"])
df[is_perc, "t_CC_SD"]   <- sd_asin(df[is_perc, "CC_mean"], df[is_perc, "CC_SD"])

df[is_perc, "t_EC_mean"] <- mean_asin(df[is_perc, "EC_mean"])
df[is_perc, "t_EC_SD"]   <- sd_asin(df[is_perc, "EC_mean"], df[is_perc, "EC_SD"])

df[is_perc, "t_CS_mean"] <- mean_asin(df[is_perc, "CS_mean"])
df[is_perc, "t_CS_SD"]   <- sd_asin(df[is_perc, "CS_mean"], df[is_perc, "CS_SD"])

df[is_perc, "t_ES_mean"] <- mean_asin(df[is_perc, "ES_mean"])
df[is_perc, "t_ES_SD"]   <- sd_asin(df[is_perc, "ES_mean"], df[is_perc, "ES_SD"])
```

## Computing effect sizes

The 2×2 design yields three quantities of interest:

1. **Main effect of enrichment**: does enrichment improve cognition, averaged across both stress conditions?
2. **Main effect of stress**:  does stress impair cognition, averaged across both enrichment conditions?
3. **Interaction**:  does the effect of enrichment differ between stressed and unstressed animals?

### Main effects

The `lnRR_main()` computes the log response ratio for a main effect. In `lnRR_main()`, the main effect is computed for "factor A" (see [documentation](https://fdecunta.github.io/minter/reference/lnRR_main.html)). 

To compute the log response ratio of enrichment, factor A is enrichment (EC) and factor B is stress (CS). Factor AB is always the interaction:


```{r}
df <- lnRR_main(
    data = df,
    col_names = c("yi_E", "vi_E"),
    Ctrl_mean = "t_CC_mean", Ctrl_sd = "t_CC_SD", Ctrl_n = "CC_n",
    A_mean    = "t_EC_mean", A_sd    = "t_EC_SD", A_n    = "EC_n",
    B_mean    = "t_CS_mean", B_sd    = "t_CS_SD", B_n    = "CS_n",
    AB_mean   = "t_ES_mean", AB_sd   = "t_ES_SD", AB_n   = "ES_n"
)
```

To obtain the stress main effect, I swap the A and B arguments:

```{r}
df <- lnRR_main(
    data = df,
    col_names = c("yi_S", "vi_S"),
    Ctrl_mean = "t_CC_mean", Ctrl_sd = "t_CC_SD", Ctrl_n = "CC_n",
    A_mean    = "t_CS_mean", A_sd    = "t_CS_SD", A_n    = "CS_n",
    B_mean    = "t_EC_mean", B_sd    = "t_EC_SD", B_n    = "EC_n",
    AB_mean   = "t_ES_mean", AB_sd   = "t_ES_SD", AB_n   = "ES_n"
)
```

The new columns `yi_E` / `vi_E` and `yi_S` / `vi_S` contain the effect sizes and their sampling variances:

```{r}
head(df[, c("ES_ID", "yi_E", "vi_E", "yi_S", "vi_S")])
```

### Interaction

The function `lnRR_inter()` computes the log response ratio of the interaction. Here the order of factors A and B is irrelevant:

```{r}
df <- lnRR_inter(
    data = df,
    col_names = c("yi_ES", "vi_ES"),
    Ctrl_mean = "t_CC_mean", Ctrl_sd = "t_CC_SD", Ctrl_n = "CC_n",
    A_mean    = "t_EC_mean", A_sd    = "t_EC_SD", A_n    = "EC_n",
    B_mean    = "t_CS_mean", B_sd    = "t_CS_SD", B_n    = "CS_n",
    AB_mean   = "t_ES_mean", AB_sd   = "t_ES_SD", AB_n   = "ES_n"
)

head(df[, c("ES_ID", "yi_ES", "vi_ES")])
```

## Aligning effect size directions

Depending on the response variable, a positive effect size can indicate better or worse performance. The data from Macartney et al. 2022 has the `Response_direction` column: a value of 2 means "lower is better". So I have to multiply those effect sizes by −1 so that positive `yi` consistently means better performance across all studies.

```{r}
df$yi_E_flip  <- ifelse(df$Response_direction == 2, -df$yi_E,  df$yi_E)
df$vi_E_flip  <- df$vi_E

df$yi_S_flip  <- ifelse(df$Response_direction == 2, -df$yi_S,  df$yi_S)
df$vi_S_flip  <- df$vi_S

df$yi_ES_flip <- ifelse(df$Response_direction == 2, -df$yi_ES, df$yi_ES)
df$vi_ES_flip <- df$vi_ES
```

## Variance–covariance matrix

Multiple effect sizes can come from the same study, so their sampling errors are correlated.  We account for this by constructing a variance–covariance matrix with `metafor::vcalc()`.  The `rho = 0.5` argument specifies the assumed correlation among effect sizes within the same study, and `cluster = Study_ID` groups them accordingly. See [Nakagawa et al. 2023](https://link.springer.com/article/10.1186/s13750-023-00301-6) for a gentle explanation. 

```{r}
VCV_E  <- vcalc(vi = vi_E_flip,  cluster = Study_ID, obs = ES_ID, rho = 0.5, data = df)
VCV_S  <- vcalc(vi = vi_S_flip,  cluster = Study_ID, obs = ES_ID, rho = 0.5, data = df)
VCV_ES <- vcalc(vi = vi_ES_flip, cluster = Study_ID, obs = ES_ID, rho = 0.5, data = df)
```

## Meta-analysis models

To estimate the average log response ratios, I fit multilevel meta-analytic models using `metafor::rma.mv()`, identicall to those used in the original paper:

### Main effect of enrichment

```{r}
mod_E <- rma.mv(
    yi = yi_E_flip, V = VCV_E,
    random = list(~1|Study_ID, ~1|ES_ID, ~1|Strain),
    test = "t", data = df
)
summary(mod_E)
```

### Main effect of stress

```{r}
mod_S <- rma.mv(
    yi = yi_S_flip, V = VCV_S,
    random = list(~1|Study_ID, ~1|ES_ID, ~1|Strain),
    test = "t", data = df
)
summary(mod_S)
```

### Interaction

```{r}
mod_ES <- rma.mv(
    yi = yi_ES_flip, V = VCV_ES,
    random = list(~1|Study_ID, ~1|ES_ID, ~1|Strain),
    test = "t", data = df
)
summary(mod_ES)
```

## Orchard plots

Orchard plots are a relatively new way of visualizing results from meta-analyses. They display the meta-analytic mean (point), 95% confidence interval (thick line), 95% prediction interval (thin line), and individual effect sizes scaled by their precision. For ecology data, where individual studies commonly produce various effect sizes, orchard plots are far better than classic forest plots. You can read more in the original paper by [Nakagawa et al. 2021](https://onlinelibrary.wiley.com/doi/full/10.1002/jrsm.1424).

### Enrichment

```{r, fig.width = 6, fig.height = 3, fig.alt = "Orchard plot showing the meta-analytic mean log response ratio for the enrichment effect across studies"}
orchard_plot(mod_E, xlab = "lnRR", alpha = 0.4, group = "Study_ID")
```

### Stress

```{r, fig.width = 6, fig.height = 3, fig.alt = "Orchard plot showing the meta-analytic mean log response ratio for the stress effect across studies"}
orchard_plot(mod_S, xlab = "lnRR", alpha = 0.4, group = "Study_ID")
```

### Interaction

```{r, fig.width = 6, fig.height = 3, fig.alt = "Orchard plot showing the meta-analytic mean log response ratio for the interaction effect"}
orchard_plot(mod_ES, xlab = "lnRR", alpha = 0.4, group = "Study_ID")
```

## Combined plot

To compare all three effects on a single figure, I have to extract the estimates from each model with `orchaRd::mod_results()`, merge them with `submerge()`, and call `orchard_plot()` once. This is not the prettiest way, but it works:

```{r}
mod_list <- list(mod_E, mod_S, mod_ES)
mod_res  <- lapply(mod_list, function(x) mod_results(x, group = "Study_ID"))

merged <- submerge(mod_res[[3]], mod_res[[2]], mod_res[[1]], mix = TRUE)

merged$mod_table$name <- factor(merged$mod_table$name,
    levels = c("Intrcpt1", "Intrcpt2", "Intrcpt3"),
    labels = rev(c("Enrichment", "Stress", "Interaction")))

merged$data$moderator <- factor(merged$data$moderator,
    levels = c("Intrcpt1", "Intrcpt2", "Intrcpt3"),
    labels = rev(c("Enrichment", "Stress", "Interaction")))
```

```{r, fig.width = 7, fig.height = 5, fig.alt = "All togheter now!"}
orchard_plot(
	merged,
	xlab = "lnRR",
	angle = 0,
	group = "Study_ID",
	alpha = 0.4,
	trunk.size = 0.8,
	branch.size = 2.8
)
```

This is the same as Figure 3A in Macartney et al. 2022. 

## References

Macartney, E. L., Lagisz, M., & Nakagawa, S. (2022). The relative benefits of environmental enrichment on learning and memory are greater when stressed: a meta-analysis of interactions in rodents. *Neuroscience & Biobehavioral Reviews*, 137, 104636. https://doi.org/10.1016/j.neubiorev.2022.104554

Nakagawa, S., Yang, Y., Macartney, E. L., Spake, R., & Lagisz, M. (2023). Quantitative evidence synthesis: a practical guide on meta-analysis, meta-regression, and publication bias tests for environmental sciences. Environmental Evidence, 12(1), 8. https://doi.org/10.1186/s13750-023-00301-6

Nakagawa, S., Lagisz, M., O'Dea, R. E., Rutkowska, J., Yang, Y., Noble, D. W., & Senior, A. M. (2021). The orchard plot: cultivating a forest plot for use in ecology, evolution, and beyond. Research Synthesis Methods, 12(1), 4-12. https://doi.org/10.1002/jrsm.1424
